---
title: "Testing"
author: "POC"
format: html
editor: visual
---

## Testing

I will use this document to work on my SIMOA code since I am having trouble with it currently in the existing format. In this way I can mess around with this code and not be worried about it impacting my original data and now that I know what I want to do for my clusters it is easier for me to manipulate the variables in a manner that will set me up for success later.

## Loading the Data

```{r}
#| label: Loading the Data
######
# In this section I will load in the dataset I want to use and name it "SIMOA" 
# so it is easier to call back on later and i know that this is my original dataset.
######

library(readr)
SIMOA <- read_csv("SIMOA Report.csv")
View(SIMOA)

```


## Eligible Particpants 

Below I am going to create a new object including only those that we can prove are ≥ 65 and are using a BZRA by answering the question (c_sp). This is to ensure they are not using other medications that may have sedating properties such as antihistamines or SSRI's and that they are older adults and fit into the research population we want.

```{r}
#| label: Eligible Participant Code

SIMOA_ELIGIBLE <- SIMOA %>% 
  # Applying the filtering based on age_cat and age
  fitler(age_cat == 1 | (age_cat == 0 & age >= 65)) %>% 
  # Applying the filtering to keep only rows where any of c_sp___1 to c_sp___14 == 1

filter(c_sp)

```

The `echo: false` option disables the printing of code (only output is displayed).
