---
title: "Testing"
author: "POC"
format: html
editor: visual
---

## Testing

I will use this document to work on my SIMOA code since I am having trouble with it currently in the existing format. In this way I can mess around with this code and not be worried about it impacting my original data and now that I know what I want to do for my clusters it is easier for me to manipulate the variables in a manner that will set me up for success later.

## Loading the Data

The section I will use to lead the dataset that I will use for the analysis.

```{r}
#| label: Loading the Data
######
# In this section I will load in the dataset I want to use and name it "SIMOA" 
# so it is easier to call back on later and i know that this is my original dataset.
######

library(readr)
SIMOA <- read_csv("SIMOA Report.csv")
View(SIMOA)

```


## Eligible Particpants

The section where I have set out the inclusion criteria to remove people from the dataset that do not meet our criteria.

```{r}
#| label: Eligible Participants
######
# In this section I will filter out those who have indicated they are <65 or that have not answered   # yes to the question about age category or not answered either question. I will also filter out those # who did not select one of the 14 BZRAs listed because we do not want the results to be affected by other sedating medications such as antihistamines or SSRI's.
######

library(tidyverse)

# Original count
n_original <- nrow(SIMOA)

# After age filtering
SIMOA_age_filtered <- SIMOA %>%
  filter(age_cat == 1 | (age_cat == 0 & age >= 65))
n_after_age <- nrow(SIMOA_age_filtered)

# After both filters
SIMOA <- SIMOA_age_filtered %>%
  filter(rowSums(select(., starts_with("c_sp___"))[, 1:14] == 1, na.rm = TRUE) > 0)
n_final <- nrow(SIMOA)

# Report results
cat("Original sample:", n_original, "\n")
cat("After age filtering:", n_after_age, "(removed:", n_original - n_after_age, ")\n")
cat("After c_sp filtering:", n_final, "(removed:", n_after_age - n_final, ")\n")
cat("Total removed:", n_original - n_final, "\n")
```


## Group Comparison


